Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-12-04 ms/spaceness.tex
minor edit to appendix
all the way through

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), \revpoint{AE}{1}
much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations,al2019estimating}\llabel{eemscitation}. For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

@@

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.

@@

To check that our model produces reasonable results, we compared its output to that of a stepping-stone model implemented in msprime \citep{Kelleher2016}. These results are discussed in detail in Appendix 1, but in general we find that our model produces many of patterns generated by stepping stone models while avoiding some artifacts associated with discretization of the landscape.

@@

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes. The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

@@

Last, we used a subset of summary statistics to compare how our continuous model performs relative to a reverse-time stepping stone model (Appendix 1).

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then "completing" the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.  \llabel{runtimesxn}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sfs_w_sumstats_loglog.pdf}
\caption{Log-scaled site frequency spectrum (A) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot. Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. \llabel{ibdnesxn} In general methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events, but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short. Combined with our analysis of IBS tract length variation (Figure \ref{fig:ibs}) this suggests that haplotype-based methods are likely to be affected by similar biases.

@@

We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}. In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$. If these models are good approximations of the continuous case we expect that results will converge as $n \to \infty$, so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}). To compare with continuous models we first distributed the the same "effective" number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations). We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma \approx 4m(W/n)$. We ran 500 simulations per $n$ while sampling $\sigma$ from $U(0.2,4)$. We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{figures/grid_ss_curves_edit.pdf}
\caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side. The black "infinite" points are from our forward-time continuous space model. Inter-deme migration rates are related to $\sigma$ as described above.}
\label{fig:grid_stats}
\end{figure*}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
\label{fig:wm_v_dd} 
\end{figure*}
\FloatBarrier

>>>>

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), \revpoint{AE}{1}
much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}. \revpoint{4}{3}
For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

@@

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates. \revpoint{4}{8}
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.

@@

We also compared our model's output to a commonly-used approximation of continuous space,
the stepping-stone model, which we simulated with \texttt{msprime} \citep{Kelleher2016}.
These results are discussed in detail in the Appendix,
but in general we find that the demographic structure of a stepping-stone model
can depend strongly on the chosen discretization,
and some artifacts of discretization do not seem to dissappear in the limit of a fine grid.
For many summary statistics, finer discretizations (we used a $50 \times 50$ grid)
produced similar results to the continuous model,
but this was not true for others (e.g., $F_{IS}$ and Tajima's $D$),
which differed from the continuous model \emph{more} at finer discretizations.

@@

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
defined to be the set of distances between adjacent sites that differ between the two chromosomes
of each diploid. \revpoint{1}{1}
The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/cjbattey/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

@@

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented running simulations at neighborhood sizes $>1000$ to coalescence is likely impractical, though running these models for more limited timescales and then ``completing'' the simulation by starting a reverse-time simulation from the resulting tree sequence in msprime is possible.  \llabel{runtimesxn}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sfs_w_sumstats_loglog.pdf}
\caption{Site frequency spectrum (A; note axes are log-scaled) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

We found that SMC++ was more robust to the effects of space than stairwayplot, underestimating recent populations by roughly half in the worst time periods rather than nearly 10-fold as with stairwayplot. Though this degree of variation in population size is certainly meaningful in an ecological context, it is relatively minor in population genetic terms. \llabel{ibdnesxn}
Methods directly assessing haplotype structure in phased data (for example, \citet{browning2015accurate}) are thought to provide increased resolution for recent demographic events,
but in this case the error we observed was essentially an accurate reflection of underlying genealogies in which terminal branches are anomalously short. Combined with our analysis of IBS tract length variation (Figure \ref{fig:ibs}) this suggests that haplotype-based methods are likely to be affected by similar biases.

@@

We also compared our model results to a regular grid of discrete populations,
which is commonly used as an approximation of continuous geography.
An important reason that this approximation is often made
is that it allows more efficient, coalescent simulations;
we implemented these using \texttt{msprime} \citep{Kelleher2016}.
In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$.
If these models are good approximations of the continuous case we expect that results
will converge as $n \to \infty$
(while scaling $m$ appropriately and keeping total population size fixed),
so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}).
To compare with continuous models we first distributed the the same ``effective'' number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations).
We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma = 4m(W/n)$.
This means that $m$ in different simulations with the same $\sigma$ scales with $\sqrt{n}$.
We ran 500 simulations for each value of $n$ while sampling $\sigma$ from $U(0.2,4)$. We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{figures/grid_ss_curves_edit.pdf}
\caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side. The black ``infinite'' points are from our forward-time continuous space model. Inter-deme migration rates are related to $\sigma$ as described above.}
\label{fig:grid_stats}
\end{figure*}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the ``Pain in the Torus'' described by \citet{Felsenstein1975}.}
\label{fig:wm_v_dd} 
\end{figure*}
\FloatBarrier
